Switch equivalence and weak prime domains for left-linear DPO rewriting systems

### 2025-04-17 full_version/permutations.tex
switch

\end{proof}

@@

\begin{lemma}\label{lem:colim}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ a left-linear DPO-rewriting system over it. The following properties hold true:
	\begin{enumerate}
		\item  if $\der{D}$ is a derivation from $G$ to $H$, then the diagram $\Delta(\der{D})$ has a colimit $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ such that $\iota_H$ belongs to $\mathcal{M}$.
		\item if $\der{D}$ is the concatenation $\der{D}_1\cdot \der{D}_2$ of two derivations $\der{D}_1=\{\dder{D}_{1,i}\}_{i=0}^{n_1}$ between $G$ and $H$ and $\der{D}_2=\{\dder{D}_{2,j}\}_{j=0}^{n_2}$ between $H$ and $T$,  then the colimiting cocone $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ exists too and there is a pushout square
		\[\xymatrix{H\ar[r]^-{\iota_{2, H}} \ar@{>->}[d]_-{\iota_{1, H}} & \tproi{D}{2} \ar[d]^{p_2}\\  \tproi{D}{1} \ar[r]_{p_1}& \tpro{D}}\]
		where $(\tproi{D}{1}, \{\iota_{1, X}\}_{X\in \Delta(\der{D}_1)})$ and $(\tproi{D}{2}, \{\iota_{2, X}\}_{X\in \Delta(\der{D}_2)})$ are the colimiting cocone for $\Delta(\der{D}_1)$ and $\Delta(\der{D}_2)$, respectively.
	\end{enumerate}
\end{lemma}
\begin{remark}\label{rem:cof}
Let $I:\Deltamin(\dder{D})\to \Delta(\dder{D})$ be the inclusion functor. It is immediate to see that such functor is \emph{final} \cite{mac2013categories}. This means that for every functor $F\colon \Delta(\dder{D})\to \Y$ we have:
\begin{enumerate}
	\item if  $(C, \{c_X\}_{X\in \Deltamin(\dder{D})})$ is colimiting for $F\circ I$, then there exists a colimiting cocone $(D, \{d_X\}_{X\in \Delta(\dder{D})})$ for $F$;
	\item $(C, \{c_X\}_{X\in \Deltamin(\dder{D})})$ and $(D, \{d_X\}_{X\in \Delta(\dder{D})})$ are colimiting for, respectively, $F\circ I$ and $F$, then the canonical arrow $\phi\colon C\to D$ induced by $(D, \{d_X\}_{X\in \Deltamin(\dder{D})})$ is an isomorphism.
\end{enumerate}
\end{remark}

@@

@@

\begin{proposition}\label{rem:abscons}
	Let $(\der{D}, \alpha, \omega)$ and $(\der{D}', \alpha', \omega')$ be two decorated derivations in a left-linear rewriting system $(\X, \R)$ of length $l$. Let also $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ be an abstraction equivalence between $\der{D}$ and $\der{D}'$. If $\alpha'\circ \phi_{G_0}=\alpha$ then the the identity $\id{[0,l-1]}$ is a semi-consistent permutation. If, moreover, $\omega'\circ \phi_{G_{l}}=\omega$, then $\id{[0,l-1]}$ is a consistent one.
\end{proposition}
\begin{proof}
	For every $X\in \Deltamin(\der{D}) $, let $X'$ the codomain of $\phi_X$. Then we can define $\xi_{\id{[0,l-1]}}\colon \tpro{D}\to \lpro \der{D}'\rpro$ as the unique morphism such that, for every $X\in \Deltamin(\der{D})$, 
	\[\xi_{\id{[0,l-1]}} \circ \iota_{X}= \iota'_{X'}\circ \phi_X\]

As a first step we can notice that $\xi_{\id{[0,l-1]}}$ is an isomorphism: indeed we can define $\Theta\colon \lpro\der{D}'\rpro \to \tpro{D}$ as the unique arrow satisfying, for every $X'\in \Deltamin(\der{D}')$
\[\Theta \circ \iota'_{X'}= \iota_{X}\circ \phi^{-1}_X\]

From the following two computations we can derive that $\Theta$ is the inverse of $\xi_{\id{[0,l-1]}}$.
\[\begin{split}
	\Theta \circ \xi_{\id{[0,l-1]}}\circ \iota_{X}&=\Theta \circ\iota'_{X'}\circ \phi_X\\&=\iota_{X}\circ \phi^{-1}_X\circ \phi_X\\&=\iota_X 
\end{split}\qquad \begin{split}
\xi_{\id{[0,l-1]}}\circ \Theta \circ \iota'_{X'}&=\xi_{\id{[0,l-1]}}\circ \iota_X\circ \phi^{-1}_{X}\\&=\iota'_{X'}\circ \phi_X\circ \phi^{-1}_{X}\\&=\iota'_{X'}
\end{split}\]

To conclude, let us observe that all the internal subdiagrams of the following diagrams commutes, guaranteeing the commutativity of the full ones.
\[\xymatrix@C=30pt{\pi(G_0)\ar[r]^{\alpha} \ar[d]_{\alpha'} & G_0 \ar[dl]^{\phi_{G_0}} \ar[r]^{\iota_{G_0}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}}\\ G'_0 \ar[rr]_{\iota'_{G'_0}} & &\lpro \der{D}' \rpro}\]
\[\xymatrix@C=30pt{L_i \ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[dl]^{\phi_{G_i}} \ar[r]^{\iota_{G_i}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}} & R_i \ar[r]^{h_i} \ar[d]_{h'_{i}}& G_{i+1} \ar[dl]^{\phi_{G_{i+1}}}\ar[r]^{\iota_{G_{i+1}}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}} \\G'_{i} \ar[rr]_{\iota'_{G'_{i}}}&& \lpro \der{D}' \rpro& G'_{i+1} \ar[rr]_{\iota'_{G'_{i+1}}}&& \lpro \der{D}' \rpro}\]

>>>>

	Thus we conclude.
\end{proof}

@@

\begin{lemma}\label{lem:colim}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ a left-linear DPO-rewriting system over it. The following properties hold true:
	\begin{enumerate}
		\item  if $\der{D}$ is a derivation from $G$ to $H$, then the diagram $\Delta(\der{D})$ has a colimit $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ such that $\iota_H$ belongs to $\mathcal{M}$.
		\item if $\der{D}$ is the concatenation $\der{D}_1\cdot \der{D}_2$ of two derivations $\der{D}_1=\{\dder{D}_{1,i}\}_{i=0}^{n_1}$ between $G$ and $H$ and $\der{D}_2=\{\dder{D}_{2,j}\}_{j=0}^{n_2}$ between $H$ and $T$,  then the colimiting cocone $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ exists too and there is a pushout square
		\[\xymatrix{H\ar[r]^-{\iota_{2, H}} \ar@{>->}[d]_-{\iota_{1, H}} & \tproi{D}{2} \ar[d]^{p_2}\\  \tproi{D}{1} \ar[r]_{p_1}& \tpro{D}}\]
		where $(\tproi{D}{1}, \{\iota_{1, X}\}_{X\in \Delta(\der{D}_1)})$ and $(\tproi{D}{2}, \{\iota_{2, X}\}_{X\in \Delta(\der{D}_2)})$ are the colimiting cocone for $\Delta(\der{D}_1)$ and $\Delta(\der{D}_2)$, respectively.
	\end{enumerate}
\end{lemma}
\begin{remark}\label{rem:cof}
Let $I:\Deltamin(\dder{D})\to \Delta(\dder{D})$ be the inclusion functor. It is immediate to see that such functor is \emph{final} \cite{mac2013categories}. This means that for every functor $F\colon \Delta(\dder{D})\to \Y$ we have:
\begin{enumerate}
	\item if  $(C, \{c_X\}_{X\in \Deltamin(\dder{D})})$ is a colimiting cocone for $F\circ I$, then there also exists a colimiting cocone $(D, \{d_X\}_{X\in \Delta(\dder{D})})$ for $F$;
	\item $(C, \{c_X\}_{X\in \Deltamin(\dder{D})})$ and $(D, \{d_X\}_{X\in \Delta(\dder{D})})$ are colimiting for, respectively, $F\circ I$ and $F$, then the canonical arrow $\phi\colon C\to D$ induced by $(D, \{d_X\}_{X\in \Deltamin(\dder{D})})$ is an isomorphism.
\end{enumerate}
\end{remark}

@@

\begin{remark}\label{rem:abscol}
Another useful property is that the abstractly equivalent derivations yield isomorphic colimits. Indeed, let $\{\phi_X\}$ be an abstraction equivalence between $\der{D}$ and $\der{E}$ and for every $X\in \Deltamin(\der{D})$ denote by $X'$ the codomain of $\phi_X$. We can define $\Phi \colon \tpro{D}\to \lpro \der{E}\rpro$ and $\Theta \colon \lpro\der{E}\rpro \to \tpro{D}$ as the unique morphisms such that, for every $X\in \Deltamin(\der{D})$:
\[\Phi \circ \iota_{\der{D},X}= \iota_{\der{E}, X'}\circ \phi_X \qquad \Theta \circ \iota_{\der{E}, X'}= \iota_{\der{D}, X}\circ \phi^{-1}_X \]

If we compute we get:.
\[\begin{split}
	\Theta \circ \Phi \circ \iota_{\der{D}, X}&=\Theta \circ\iota_{\der{E}, X'}\circ \phi_X\\&=\iota_{\der{D}, X}\circ \phi^{-1}_X\circ \phi_X\\&=\iota_{\der{D}, X} 
\end{split}\qquad \begin{split}
	\Phi \circ \Theta \circ \iota_{\der{E}, X'}&=\Phi \circ \iota_{\der{D}, X}\circ \phi^{-1}_{X}\\&=\iota_{, \der{E}, X'}\circ \phi_X\circ \phi^{-1}_{X}\\&=\iota_{\der{E},X'}
\end{split}\]

Therefore, $\Phi$ is an isomorphism with inverse $\Theta$.
\end{remark}

@@

\begin{proposition}\label{rem:abscons}
	Let $(\der{D}, \alpha, \omega)$ and $(\der{D}', \alpha', \omega')$ be two decorated derivations in a left-linear rewriting system $(\X, \R)$ of length $l$. Let also $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ be an abstraction equivalence between $\der{D}$ and $\der{D}'$. If $\alpha'\circ \phi_{G_0}=\alpha$ then the the identity $\id{[0,l-1]}$ is a semi-consistent permutation. If, moreover, $\omega'\circ \phi_{G_{l}}=\omega$, then $\id{[0,l-1]}$ is a consistent one.
\end{proposition}
\begin{proof}
By \Cref{rem:abscol} we know that there is an isomorphism  $\xi_{\id{[0, l-1]}}\colon \tpro{D}\to \lpro \der{D}'\rpro$.  We can notice that all the internal subdiagrams of the following diagrams commutes, guaranteeing the commutativity of the full ones.
\[\xymatrix@C=30pt{\pi(G_0)\ar[r]^{\alpha} \ar[d]_{\alpha'} & G_0 \ar[dl]^{\phi_{G_0}} \ar[r]^{\iota_{G_0}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}}\\ G'_0 \ar[rr]_{\iota'_{G'_0}} & &\lpro \der{D}' \rpro}\]
\[\xymatrix@C=30pt{L_i \ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[dl]^{\phi_{G_i}} \ar[r]^{\iota_{G_i}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}} & R_i \ar[r]^{h_i} \ar[d]_{h'_{i}}& G_{i+1} \ar[dl]^{\phi_{G_{i+1}}}\ar[r]^{\iota_{G_{i+1}}} &\tpro{D} \ar[d]^{\xi_{\id{[0, l-1]}}} \\G'_{i} \ar[rr]_{\iota'_{G'_{i}}}&& \lpro \der{D}' \rpro& G'_{i+1} \ar[rr]_{\iota'_{G'_{i+1}}}&& \lpro \der{D}' \rpro}\]
